Adversarial Training against Systematic Uncertainty

### 2016-09-02 tex/paper.tex
unify categorical and continuous parameters

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec in faucibus
lacus. Curabitur in viverra neque. Sed nec urna a sem blandit viverra vel a
eros. Nam a ultrices leo, a aliquet diam. Proin cursus scelerisque leo, ut
ullamcorper ante auctor et. Nulla odio nisl, molestie eget consectetur quis,
eleifend id ligula. Cras lobortis egestas justo ac sagittis. In at ligula est.
Nulla vel eros sapien. Sed bibendum metus quis dignissim scelerisque. Nulla arcu
lectus, ornare nec vehicula quis, lobortis a mi. Cras scelerisque mauris quam,
nec viverra ex convallis sit amet. Ut tellus turpis, lobortis sit amet finibus
sit amet, maximus ac est.

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary classifier $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto {\cal Z}$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$.
Assuming that ${\cal Z}$ defines a finite family of nuisance values $z_l$ (for $l=1, \dots, |{\cal Z}|$),
this classifier takes as input realizations of $f(X; \theta_f)$, for the current value $\theta_f$ of $f$ parameters,
and produces as output probability estimates $r(f(X; \theta_f); \theta_r)_l = \hat{p}(z_l| f(X; \theta_f)) $
that $f(X; \theta_f)$ is generated from the nuisance value $z_l$. If $p(f(X; \theta_f)|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)|z)$ is invariant with $z$, as we require, then $r$
should perform poorly and be close to random guessing. Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
The adversarial training procedure to obtain $(\smash{\hat\theta_f},
\smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of $f$ being a classifier
and of the cross-entropy loss for both ${\cal L}_f$ and ${\cal L}_r$. The algorithm
consists in using stochastic gradient descent alternatively to optimize
Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

\begin{algorithm}[t]
\caption{Adversarial training of a classifier $f$ against an adversary $r$.\\
{\it Inputs:} training data $\{ x_i, y_i, z_i \}_{i=1}^N$\\
{\it Outputs:} $\smash{\hat\theta_f}, \smash{\hat\theta_r}$\\
{\it Hyper-parameters:} Number $T$ of training iterations, Number $K$ of gradient steps to update $r$.}
\label{alg:adversarial-training}
\begin{algorithmic}[1]
    \For{$t=1$ to $T$}
        \For{$k=1$ to $K$} \Comment{Update $r$}
            \State{Sample minibatch $\{x_m, z_m \}_{m=1}^M$} of size $M$;
            \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
            $$\nabla_{\theta_r} \sum_{m=1}^M \left[\sum_{z_l \in {\cal Z}} 1(z_m = z_l)\log r(f(x_m;\theta_f);\theta_r)_l \right];$$}
        \EndFor
        \State{Sample minibatch $\{x_m, y_m, z_m \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
        \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
        $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\sum_{y_c \in {\cal Y}} 1(y_m = y_c) \log f(x_m;\theta_f)_c  +\sum_{z_l \in {\cal Z}} 1(z_m = z_l)\log r(f(x_m;\theta_f);\theta_r)_l \right];$$}
    \EndFor
\end{algorithmic}
\end{algorithm}

\glnote{Extension to continuous nuisance parameters: have the adversary $r$
models the conditional distribution $p(z|f(X;\theta_f))$, e.g. with a network
whose 2 outputs are the parameters of a truncated gaussian and optimized such that
$p(z|f(X;\theta_f)) \approx N(\mu(f(X;\theta_f);\theta_r), \sigma(f(X;\theta_f);
\theta_r)$. The associated loss should be the negative log-likelihood (= cross-entropy), which is maximal when the
approximation is perfect and minimal when the truncated gaussian degenerates to
a uniform (when $\sigma$ is large), in which case we should have the guarantee
that $f$ is pivotal.}

\glnote{This section and theoretical results should be rewritten in such a way
that categorical and continuous nuisance values are solved under the same
principled approach.}

@@

\section{Theoretical results}

In this section, we show that in the setting of
Algorithm~\ref{alg:adversarial-training}, the procedure converges to a
classifier $f$ which is a pivotal quantity in the sense of
Eqn.~\ref{eqn:criterion}. Results below are derived in a non-parametric setting,
by assuming that both $f$ and $r$ have enough capacity. To simplify the
presentation, we also assume the uniform prior $p(z) =
\frac{1}{|{\cal Z}|}$ for all $z \in {\cal Z}$, e.g. by having the same
number of training samples for each modality $z$ of the nuisance
parameter.

\begin{proposition}\label{prop:1}
Let $\theta_f$ be fixed and $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$. If $r(f(X;\theta_f) ; \hat\theta_r )_l = \frac{1}{|{\cal Z}|}$
for all $z_l$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
Let us first recall that the cross-entropy for distributions $p$ and $q$ is
minimized when $p=q$. For ${\cal L}_r$ defined as the cross-entropy between the
true conditional distribution of the nuisance $p_{Z|f(X;\theta_f)}$ and
the approximate conditional distribution of the nuisance
$p_{r(f(X;\theta_f);\theta_r)|f(X)}$, the optimal parameters $\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} L_r(\theta_f,
\theta_r)$ are therefore such that $p_{r(f(X;\theta_f);\hat\theta_r)|f(X)} = p_{Z|f(X;\theta_f)}$.

In other words, for all $z_l \in {\cal Z}$, we have $r(f(X;\theta_f);\smash{\hat\theta}_r)_l = p(z_l|f(X;\theta_f))$.
By assumption, $r(f(X;\theta_f);\smash{\hat\theta}_r)_l = \frac{1}{|{\cal Z}|}$,
and therefore $p(z_l|f(X;\theta_f)) = \frac{1}{|{\cal Z}|}$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z_l) &= \frac{ p(z_l|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z_l)} \\
                         &= \frac{ \tfrac{1}{|{\cal Z}|} p(f(X;\theta_f)) } { \tfrac{1}{|{\cal Z}|} } \\
                         &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z_l \in {\cal Z}$ and implies that $f$ is a pivotal quantity.
\end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = H(p_{Y|X}) - \log |{\cal Z}|$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier  and a pivotal quantity.
\end{proposition}

\begin{proof}
For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} L_r(\theta_f,
\theta_r)$, in which case $p_{r(f(X;\theta_f);\hat\theta_r)|f(X)} = p_{Z|f(X;\theta_f)}$ and $L_r$ reduces to the entropy
$H(p_{Z|f(X;\theta_f)})$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = L_f(\theta_f) -
H(p_{Z|f(X;\theta_f)}).$$  In particular, we have the lower bound $H(p_{Y|X}) -
\log |{\cal Z}| \leq L_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$ where the equality
holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f} E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ corresponds to
    the parameters of an optimal classifier, in which case the log-loss $L_f$
    reduces to its minimum value $H(p_{Y|X})$,
    \item all
   outcomes of $Z|f(X;\smash{\hat\theta}_f)$ are equally likely, in which case
   $p(z_l|f(X;\smash{\hat\theta}_f)) = \frac{1}{|{\cal Z}|}$ for all $z_l \in {\cal
   Z}$ and $H(p_{Z|f(X;\smash{\hat\theta}_f)}) =
   -\sum_{z_l \in {\cal Z}} p(z_l|f(X;\smash{\hat\theta}_f)) \log  p(z_l|f(X;\smash{\hat\theta}_f)) =  -\sum_{z_l \in {\cal Z}} \frac{1}{|{\cal Z}|} \log \frac{1}{|{\cal Z}|} = \log |{\cal Z}|$.
\end{itemize}
Accordingly, the second condition implies that $r(f(X;\smash{\hat\theta}_f) ;
\smash{\hat\theta}_r )_l = \frac{1}{|{\cal Z}|}$ and therefore that at this
point, because of Proposition~\ref{prop:1}, the optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal quantity.
\end{proof}

>>>>

Lorem ipsum dolor sit amet, consectetur adipiscing elit...

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r := p_{\theta_r}(z |
f(X;\theta_f))$ of parameters $\theta_r$ and associated loss ${\cal
L}_r(\theta_f, \theta_r)$. This model takes  as input realizations of $f(X;
\theta_f)$, for the current value $\theta_f$ of $f$ parameters, and produces as
output probability estimates $p_{\theta_r}(z | f(X;\theta_f))$ that $f(X;
\theta_f)$ is generated from the nuisance value $z$. Intuitively, if $p(f(X; \theta_f)|z)$
varies with $z$, then the corresponding correlation can be captured by $r$. By
contrast, if $p(f(X; \theta_f)|z)$ is invariant with $z$, as we require, then
$r$ should perform poorly and be close to random guessing. Training $f$ such
that it additionally minimizes the performance of $r$ therefore acts as a
regularization towards Eqn.~\ref{eqn:criterion}.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Section~\ref{sec:theory}, ${\cal L}_f$ (resp. ${\cal L}_r$) is set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ (resp. of $Z|f(X;\theta_f)$ under
$r$). The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

\begin{algorithm}[t]
\caption{Adversarial training of a classifier $f$ against an adversary $r$.\\
{\it Inputs:} training data $\{ x_i, y_i, z_i \}_{i=1}^N$\\
{\it Outputs:} $\smash{\hat\theta_f}, \smash{\hat\theta_r}$\\
{\it Hyper-parameters:} Number $T$ of training iterations, Number $K$ of gradient steps to update $r$.}
\label{alg:adversarial-training}
\begin{algorithmic}[1]
    \For{$t=1$ to $T$}
        \For{$k=1$ to $K$} \Comment{Update $r$}
            \State{Sample minibatch $\{x_m, z_m \}_{m=1}^M$} of size $M$;
            \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
            $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|f(x_m;\theta_f))  ;$$}
        \EndFor
        \State{Sample minibatch $\{x_m, y_m, z_m \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
        \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
        $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|f(x_m;\theta_f))  \right],$$
        \indent where $p_{\theta_f}(y_m|x_m)$ denotes $1(y_m=0)(1-f(x_m;\theta_f)) + 1(y_m=1)f(x_m;\theta_f))$;}
    \EndFor
\end{algorithmic}
\end{algorithm}

@@

\section{Theoretical results}
\label{sec:theory}

In this section, we show that in the setting of
Algorithm~\ref{alg:adversarial-training} where ${\cal L}_r$ is set to expected
value of the negative log-likelihood of $Z|f(X;\theta_f)$ under $r$, the
procedure converges to a classifier $f$ which is a pivotal quantity in the sense
of Eqn.~\ref{eqn:criterion}. Propositions below are derived in a non-parametric
setting, by assuming that both $f$ and $r$ have enough capacity. Results hold
for a nuisance parameter $Z$ taking either categorical values or continuous values within a
bounded support.  By abuse of notation, $H(p_Z)$ denotes the differential
entropy in this latter case. Finally, by construction, we assume the uniform
prior $p(z)$ over the nuisance values $z \in {\cal Z}$ of the parameter.

\begin{proposition}\label{prop:1}
Let $\theta_f$ be fixed and $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$. If $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
The optimal parameters $\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} L_r(\theta_f,
\theta_r)$ are such that $p_{\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)} = p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and implies that $f$ is a pivotal quantity.
\end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = H(p_{Y|X}) - H(p_Z)$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier  and a pivotal quantity.
\end{proposition}

\begin{proof}
For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} L_r(\theta_f,
\theta_r)$, in which case $p_{\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$ and $L_r$ reduces to the entropy
$H(p_{Z|f(X;\theta_f)})$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = L_f(\theta_f) -
H(p_{Z|f(X;\theta_f)}).$$  In particular, we have the lower bound $H(p_{Y|X}) -
H(p_Z) \leq L_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$ where the equality
holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f} E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ corresponds to
    the parameters of an optimal classifier, in which case the expected negative log-likelihood $L_f$ of $Y|X$
    reduces to its minimum value $H(p_{Y|X})$,
    \item all
   outcomes of $Z|f(X;\smash{\hat\theta}_f)$ are equally likely, in which case
   $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal
   Z}$ since we assume a uniform prior by construction. Note that in the continuous case, this bound is also realized
   by the uniform distribution when the support of $Z$ is bounded.
\end{itemize}
Accordingly, the second condition implies that $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$ and therefore that at this
point, because of Proposition~\ref{prop:1}, the optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal quantity.
\end{proof}
